---
title: "About"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{About}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "",
  R.options = list(
    cli.num_colors = 1,
    cli.hyperlink = FALSE,
    crayon.enabled = FALSE,
    width = 80
  )
)
# Console colour carries no meaning on a rendered page. pkgdown turns it on for
# its own build, and the escape sequences then reach the reader as literal text,
# so colour is switched off here for a plain vignette render and a site build
# alike. The fixed width keeps printed output inside the documentation column.
```

## Citing lexsync

If lexsync contributes to published work, please cite it. The reference below
points at the DOI that [CRAN](https://CRAN.R-project.org/package=lexsync)
assigned to the package.

> Bernabeu, P. (2026). lexsync: Lexical optimisation and hardware-timed
> experiment generation. R package version
> `r packageVersion("lexsync")`. https://doi.org/10.32614/CRAN.package.lexsync

```{r bibtex, echo = FALSE, results = "asis"}
# Build the BibTeX entry from the installed version so it never drifts, then
# render it with a copy button and a download link. The button uses the
# browser clipboard API; the link is a self-contained data URI, so neither
# depends on a static file being shipped alongside the site.
ver <- as.character(utils::packageVersion("lexsync"))
bib <- paste(
  "@Manual{lexsync,",
  "  title  = {{lexsync}: Lexical optimisation and hardware-timed experiment generation},",
  "  author = {Pablo Bernabeu},",
  "  year   = {2026},",
  sprintf("  note   = {R package version %s},", ver),
  "  doi    = {10.32614/CRAN.package.lexsync},",
  "  url    = {https://CRAN.R-project.org/package=lexsync},",
  "}",
  sep = "\n"
)
esc <- function(x) {
  x <- gsub("&", "&amp;", x, fixed = TRUE)
  x <- gsub("<", "&lt;", x, fixed = TRUE)
  gsub(">", "&gt;", x, fixed = TRUE)
}
uri <- paste0(
  "data:application/x-bibtex;charset=utf-8,",
  utils::URLencode(bib, reserved = TRUE)
)
cat(sprintf(
'<div class="citation-bibtex">
<pre id="lexsync-bibtex"><code>%s</code></pre>
<p class="citation-bibtex-actions">
<button type="button" class="btn btn-primary btn-sm" onclick="lexsyncCopyBibtex(this)">Copy BibTeX</button>
<a class="btn btn-outline-primary btn-sm" download="lexsync.bib" href="%s">Download .bib</a>
</p>
</div>
<script>
function lexsyncCopyBibtex(btn) {
  var code = document.getElementById("lexsync-bibtex");
  navigator.clipboard.writeText(code.innerText).then(function () {
    var label = btn.textContent;
    btn.textContent = "Copied";
    setTimeout(function () { btn.textContent = label; }, 1500);
  });
}
</script>
', esc(bib), uri))
```

R users can also retrieve this citation with `citation("lexsync")`, although
the 0.1.0 release on CRAN, built before the DOI existed, prints the GitHub
address and no DOI. The repository carries a machine-readable
[`CITATION.cff`](https://github.com/pablobernabeu/lexsync/blob/main/CITATION.cff)
as well, which GitHub turns into a ready-made citation through its 'Cite this
repository' button and which reference managers can import.

A manuscript describing lexsync is in preparation, under the title *lexsync: A
cross-platform pipeline for multidimensional lexical optimisation and
hardware-timed experiment generation*. It is unpublished and has no venue as
yet. Once it is accepted, `CITATION.cff` will name it as the preferred citation,
and until then the software reference above is the one to use.

## Citing the corpus

The software citation covers the tool, not the data your items came from. The
corpora are third-party work with terms of their own, and every one is credited,
with its licence and its retrieval date, in
[`corpora/ATTRIBUTION.md`](https://github.com/pablobernabeu/lexsync/blob/main/corpora/ATTRIBUTION.md).
Each run's materials datasheet records the source file the run read and its
SHA-256, so the corpus behind a published stimulus set stays identifiable long
after the run, and a reader can check that the file they hold is the file the
selection was made from.

## The developer

[Pablo Bernabeu](https://pablobernabeu.github.io) is a researcher in the
Department of Education at the University of Oxford, with hands-on experience of
behavioural experiments, EEG, corpus analysis, computational modelling and
statistics. He develops open, reproducible research software in R and Python,
and is a Fellow of the Software Sustainability Institute. His
[ORCID record](https://orcid.org/0000-0003-1083-2460) lists his other work.

lexsync has a feature-parity twin in Python, documented at
[its own site](https://pablobernabeu.github.io/lexsync/python/). The two
packages are built from one repository and released in step under one version,
and under the deterministic matching methods they select byte-identical stimuli
from the same lexicon and design. A group can therefore work in whichever
language suits it without the materials diverging.

## Licence

The source code is released under the
[MIT licence](https://pablobernabeu.github.io/lexsync/r/LICENSE.html). The
corpus derivatives bundled with the package are not covered by it. They inherit
share-alike terms from the corpora they were built from, so they are distributed
under CC BY-SA 4.0 instead, for the reasons
[`LICENSE-DATA`](https://github.com/pablobernabeu/lexsync/blob/main/LICENSE-DATA)
sets out, with the sources and retrieval dates recorded in
[`corpora/ATTRIBUTION.md`](https://github.com/pablobernabeu/lexsync/blob/main/corpora/ATTRIBUTION.md). The distinction becomes practical if you redistribute
a lexicon derived from the bundled data, since that means crediting the original
corpus authors, saying what was changed and passing the same licence on. Nothing
of the sort applies to the code you write with the package. A reader who has only
the package rather than the repository will find the same terms, and the credit
they require, in the `LICENSE.note` file that ships in the sources.

## Versioning and archival

lexsync follows semantic versioning, and the R and Python packages share one
version number so that a citation identifies the same state of both. Releases
are tagged on GitHub, and the
[changelog](https://pablobernabeu.github.io/lexsync/r/news/index.html) records
what changed in each of them.

The R package is on [CRAN](https://CRAN.R-project.org/package=lexsync), which
keeps the source of every version it publishes. A single DOI,
[10.32614/CRAN.package.lexsync](https://doi.org/10.32614/CRAN.package.lexsync),
covers all of those versions, so the reference above names the version as well.
Zenodo also archives every GitHub release, which carries both packages, under
the concept DOI [10.5281/zenodo.22906962](https://doi.org/10.5281/zenodo.22906962). That is the
identifier the Python package cites, since the CRAN DOI names the R package
alone.

## Contributing and support

Bugs and feature requests are welcome on the
[issue tracker](https://github.com/pablobernabeu/lexsync/issues), which is also
where questions about a design or a generated experiment are best raised. The
[contributing guide](https://github.com/pablobernabeu/lexsync/blob/main/.github/CONTRIBUTING.md) describes the development setup for
both engines and the parity rule that governs every change, and everyone taking
part is asked to honour the [Code of Conduct](https://github.com/pablobernabeu/lexsync/blob/main/.github/CODE_OF_CONDUCT.md).

A report worth acting on carries the design YAML that was run alongside the run
log that `run_pipeline()` wrote beside the results. Between them they pin the
inputs and every step that ran, which is usually enough to reproduce a problem
without a further round of questions. Where a corpus is involved, the materials
datasheet from the same run identifies it by path and checksum.
