---
title: "Experiments, paradigms and EEG triggers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Experiments, paradigms and EEG triggers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "",
  R.options = list(
    cli.num_colors = 1,
    cli.hyperlink = FALSE,
    crayon.enabled = FALSE,
    width = 80
  )
)
# Console colour carries no meaning on a rendered page. pkgdown turns it on for
# its own build, and the escape sequences then reach the reader as literal text,
# so colour is switched off here for a plain vignette render and a site build
# alike. The fixed width keeps printed output inside the documentation column.
```

Once a stimulus set exists, something has to present it. `lexsync` generates that
experiment from a declarative description of the trial, sparing the hand-coding
that usually follows selection. A trial is a list of events, each event says what
appears and for how long and whether a hardware trigger rides on its onset, and
each presentation target is an interpreter of that same list. The arrangement is
what lets one engine serve several paradigms and three targets: a new paradigm is
a new event list, not new backend code.

This vignette covers the event model, the paradigm registry, the three targets
and how they differ, the trigger timing that motivates the design, and how to add
a paradigm. It runs on the bundled English example lexicon and writes everything
to a temporary directory.

```{r setup}
library(lexsync)
schema <- yaml::read_yaml(
  system.file("extdata", "schema.yaml", package = "lexsync")
)
lex <- load_lexicon(
  system.file("extdata", "en_example.csv", package = "lexsync"),
  schema, language = "english"
)
pool <- build_pool(lex, list(length = c(3, 7), frequency = c(3.8, 7.0)))

design <- list(
  name = "vignette_experiment", language = "english", n_per_condition = 12,
  paradigm = "factorial",
  conditions = list(
    list(name = "high", define_by = list(frequency = c(5.2, 7.0))),
    list(name = "low",  define_by = list(frequency = c(3.8, 4.4)))
  ),
  match_on = list("length", "n_density", "old20"),
  counterbalance = list(lists = 1)
)
stim <- match_stimuli(pool, design, schema)
stim <- counterbalance(stim, design, schema)
head(stim[, c("trial", "list", "set", "condition", "word")], 3)
```

## The trial-event model

A design's trial is a list of events. Each event is a plain list whose `type`
selects the behaviour and whose remaining fields parameterise it. The factorial
paradigm's default sequence is a fixation cross, the word, a response window and
a blank.

```{r events}
events <- resolve_events(design)
str(events)
```

`resolve_events()` is the accessor the backends use, and it embodies the
precedence rule: a design's own `events` list wins, and only if there is none does
the design inherit its paradigm's default sequence. The eight event types are
`fixation`, `text`, `mask`, `blank`, `region_by_region`, `response`, `question`
and `feedback`. The first four draw something (or nothing, for `blank`) for a fixed
number of frames. `response` and `question` collect a keypress. `region_by_region`
presents a segmented sentence one region at a time, advancing on a key.
`feedback` scores the keypress just collected and reports it, and is described
under practice blocks below.

Durations are given in milliseconds, as `duration_ms`. That is the unit a
methods section reports and the unit all three targets present, so a design means
the same interval wherever it runs. A visual stimulus can still only appear on a
refresh, so the PsychoPy script measures the display's refresh rate when it starts
and converts each duration into the nearest whole number of flips: 50 ms is three
flips at 60 Hz and seven at 144 Hz. The trigger stays bound to the flip, and the
residual quantisation is reported.

A duration may also be given as `duration_frames`, which designs written before
this change still use. A frame count is converted to milliseconds when the design
loads, at `presentation.assumed_refresh_hz` in the schema (60 by default), so it
means the rate the count was written for, whatever rate the experiment then runs
at. The `timeout_ms` of a response window has always been in milliseconds,
because a response deadline is not tied to the refresh.

### Content and field references

An event's `content` is either a literal or a reference to a column of the trial
table. A reference is a single field name in braces, and it is what makes an
event list a template: one list serves every trial in the design.

```{r content}
lapply(events[1:2], function(ev) ev$content)
```

The fixation event's content is the literal `+`, which every trial shows.
The text event's content is `{word}`, so each trial shows its own row's `word`
column. `required_fields()` resolves the references into the columns a design's
items must actually carry, which is how a design's item table is validated before
anything is generated.

```{r required-fields}
required_fields(design)
required_fields(list(paradigm = "priming"))
required_fields(list(paradigm = "self_paced_reading"))
```

Presented strings are always carried as data. Each target writes a trial table and
reads the strings from it at run time, and they are never interpolated into generated
code. That is why an apostrophe in a stimulus cannot break a generated script, and
it is enforced on the way in: `load_items()` rejects control characters and
over-long values in the fields a paradigm presents.

### Triggers on an event

An event may carry a `trigger`. Three forms are accepted, and the choice of form
is the choice of what the EEG marker will mean.

```{r triggers-in-events}
priming_events <- PARADIGMS$priming$events
or_dash <- function(x) if (is.null(x)) "-" else as.character(x)
data.frame(
  type = vapply(priming_events, function(e) e$type, character(1)),
  content = vapply(
    priming_events, function(e) or_dash(e$content), character(1)
  ),
  trigger = vapply(priming_events, function(e) or_dash(e$trigger), character(1))
)
```

The literal integer 20 on the prime marks every prime with the same code, since
what is being timestamped is the prime's appearance, whichever prime it is. The
token `"condition"` on the target resolves per trial to that trial's condition
code, which is the marker an ERP analysis will epoch on. The token `"item"`
resolves to a per-item code. Events with no trigger send nothing.

The tokens are resolved by `assign_triggers()`, which computes the codes and adds
them as columns.

```{r assign-triggers}
trig <- assign_triggers(stim)
unique(trig[, c("condition", "condition_trigger")])
range(trig$item_trigger)
```

Condition codes start at 101 and count up in order of first appearance, unless
`assign_triggers()` is given the order through its `conditions` argument. The
order of first appearance in a shuffled trial list depends on the seed, so an
analysis that decodes the codes should not rely on it. `export_experiments()`
therefore passes the order of the design's `conditions` entries, and the
pipeline does the same, adding any other condition in the order the item source
lists it, such as the words before the pseudowords of a generated lexical
decision. The first condition the design lists is always 101, whatever the seed.

```{r assign-triggers-order}
unique(assign_triggers(stim, conditions = c("low", "high"))[
  , c("condition", "condition_trigger")])
```

Item codes run from 40 upwards and wrap at 200 values, so a design with more than
200 items reuses item codes. That ceiling comes from the 8-bit parallel port: a
code is one byte, and the block markers 254 and 255 are reserved. Condition codes
stay unique regardless, so the analysis that matters is unaffected.


### Timing that varies from trial to trial

A duration need not be the same on every trial. An event may instead declare a
`duration:` block, in one of two forms:

```yaml
- type: text
  content: '{prime}'
  duration: {from_column: soa_ms}            # read per trial from the items
- type: blank
  duration: {jitter: [400, 800], as: iti_ms} # drawn per trial, in milliseconds
```

The two exist for different reasons. A duration read from a column is a
manipulated variable: the stimulus-onset asynchrony of a priming study is the
lever that separates automatic from strategic processing, so it belongs in the
item table and in the analysis. A jittered duration is not manipulated at all. It
decorrelates the design matrix, as EEG and fMRI designs routinely require.

Neither draws a random number. A jittered value is a uniform integer keyed on the
seed, the column name, the list, the set and the condition, so both engines
realise the same milliseconds and a rerun reproduces them. Naming the column in
the key is what makes two jittered events draw independently, since without it
they would share a single value.

Either form writes the realised milliseconds into the stimuli table and the loop
table, because timing that varies is a variable the analysis needs, not
presentation detail. `config/design_en_priming_jitter.yaml` is a worked
example carrying both.

## The paradigm registry

A paradigm bundles a default event sequence with the fields it needs and the
recipe for counterbalancing it. `PARADIGMS` is the registry, and it is exported,
so it can be read to see exactly what a design will inherit.

```{r registry}
names(PARADIGMS)
data.frame(
  paradigm = names(PARADIGMS),
  fields = vapply(
    PARADIGMS,
    function(p) paste(p$stimulus_fields, collapse = ", "),
    character(1)
  ),
  counterbalance = vapply(
    PARADIGMS, function(p) p$counterbalance, character(1)
  ),
  n_events = vapply(PARADIGMS, function(p) length(p$events), integer(1)),
  row.names = NULL
)
```

`factorial` shows a matched word per trial. `lexical_decision` is the same shape
but presents a `target`, which is a real word or a generated pseudoword.
`priming` adds a brief prime and a mask before the target. `self_paced_reading`
replaces the word event with a region-by-region sentence and adds a comprehension
question. `categorisation` shows a category cue and then the word to judge against
it.

That last one is worth a paragraph, because what separates it from lexical
decision is not the shape of the trial but where the question lives. The cue is
a trial event, shown afresh each time, since the category varies from trial to
trial, and crossing one word with two cues is how a categorisation study
separates a property of the word from the demands of the task. A robin is a bird
quickly and an animal slowly, and only the question changed. Its `answer` field
holds the key that is correct on the trial, so scoring is a string comparison
against the recorded response with nothing to look up in whatever language the
analysis is written in. The paradigm requires the field, which means an
unscoreable categorisation experiment cannot be generated.

The `counterbalance` entry is the part that is easy to overlook and important to
get right. The factorial recipe shows every matched item and splits the matched
sets across lists. The Latin-square recipe exists because `priming`,
`self_paced_reading` and `categorisation` have a property the word paradigms do
not: the same target appears in more than one condition. Showing a participant
both the related and the unrelated version of one target would let them see the
manipulation, and showing the same word under both category cues would turn the
second presentation into a repetition-priming trial. The Latin square gives each
item exactly one condition per list and rotates it across lists, so no target
repeats within a list and conditions stay balanced.

```{r counterbalance-recipes}
table(list = stim$list, condition = stim$condition)
```

### Balanced list assignment

The factorial deal sends set 1 to list 1, set 2 to list 2 and so on. That is
reproducible, but it balances nothing: every *n*th set lands in the same list, so
a dimension that happens to vary smoothly across sets is dealt out unevenly, and
where each list goes to a different group of participants, the unevenness is
confounded with the group.

`counterbalance.optimise` searches instead for an assignment whose lists have
near-equal totals on the dimensions you name, by exchanging pairs of item sets
between lists. List sizes are preserved, since a swap trades one set for another.

```yaml
counterbalance:
  lists: 4
  optimise: true
  balance_on: [length, n_density, old20, frequency]
```

`balance_on` defaults to `match_on`, and the example widens it deliberately.
Frequency is the manipulated variable and so is not matched on, but it is
manipulated *within* a list, since every list holds both conditions. Equating the
lists on its total therefore costs the manipulation nothing and removes a
difference between the participant groups who receive different lists. Naming only
the matched dimensions leaves frequency dealt arbitrarily, and measurably so: on
the shipped design the optimiser then improves the three named dimensions and
makes frequency worse than the arbitrary deal had it. Balance what you want
equated across lists, which is usually everything.

This is a steepest descent to a local optimum, not a global search. What it
guarantees is that no single exchange would improve matters further, and the
datasheet records the imbalance before and after, so the improvement is checkable.
The objective is all-integer and ties are broken by the seeded keyed hash, which
keeps the two engines on the same assignment and stops list 1 being favoured for
being numbered first.

It is off by default, and stays off. Switching it on changes which items a
participant sees, so it has to be a deliberate design decision. A package
upgrade must never make that change for a study that is already running. It is
refused on a Latin-square design, where every item already appears in every list
and the lists are balanced on the items by construction. `balance_lists()` runs
the search alone if you want the assignment without applying it, and
`design_en_balanced_lists.yaml` is the worked example.

## Practice, fillers and feedback

Everything above treats one frame as both the materials record and the thing that
runs. That holds only while the two are the same trials, and they usually are not.
Practice exists to settle the participant into the task and is discarded before
analysis. Fillers exist to dilute the manipulation so the participant cannot guess
it, and are likewise not analysed. Both have to reach the generated experiment,
and neither belongs in the stimuli file, the descriptives or the realised control.

The pipeline therefore splits. The stimuli CSV and the reports are written from
the main rows. The PsychoPy, OpenSesame and jsPsych experiments are generated from
every presented trial. A `block` column marks which is which, and it appears only
when a design declares the blocks, so a design without them keeps exactly the
columns it had.

```yaml
practice:
  path: items/practice_en_lexdec.csv
fillers:
  path: items/fillers_en_lexdec.csv
```

Where each block goes is a methodological choice. Practice comes first, as its
own run, shuffled within itself so participants do not all meet the practice
items in one order. Fillers are interleaved with the main trials, because a
block of fillers at the end is not a filler at all: it is a second block the
participant can tell apart. They are merged in before the order is drawn, so one
deterministic shuffle mixes them through, which does renumber the main trials.
That is correct, since adding fillers changes the sequence and the stimuli file
records where each item actually appeared. Both blocks appear in every list and
neither is counterbalanced, because they carry no manipulation to rotate and
every participant should get the same practice.

Each block's item table is read with the same validation as any other, and given a
`set` range that cannot collide with the main items, which a naive read would not
manage since practice item 1 and main item 1 would both be set 1. The counts and
the tables' checksums go into the datasheet, because what the participant saw is
part of the materials even when it is not part of the analysis.

A `feedback` event scores the trial and shows the result. It reads the field named
by `answer`, compares it as a string with the key the participant pressed, and
displays `correct`, `incorrect` or `no_response` for `duration_ms`.

```yaml
- type: feedback
  answer: answer
  correct: 'Correct'
  incorrect: 'Incorrect'
  no_response: 'Too slow'
  duration_ms: 600
  blocks: [practice]
```

`blocks:` restricts an event to the named blocks, and this is its main use:
feedback teaches the mapping during practice, and would contaminate reaction times
in the task itself. The restriction has to be expressed on the event because the
event list is global to the design. Since a feedback event scores a keypress,
something before it must have collected one. A design whose feedback event has no
preceding `response` or `question` is refused when the experiment is generated.
Left to run time, that one design error would surface as three different failures,
one per target. `design_en_lexdec_blocks.yaml` puts all of this together.

## The three presentation targets

The same rendered event list feeds all three exporters, so the procedure is the
same in each. What differs is what the platform can do.

```{r export}
out <- file.path(tempdir(), "lexsync_experiment")
dir.create(out, showWarnings = FALSE)
files <- export_experiments(stim, design, schema, out)
basename(unlist(files))
```

`export_experiments()` calls `assign_triggers()` first, then each exporter, and
returns the three experiment scripts. The directory holds more than that, and the
difference is the interesting part.

```{r export-dir}
list.files(out)
```

PsychoPy and OpenSesame each got their own trial table and read it at run time,
which is why two CSVs appear alongside the three scripts. The jsPsych target has
no second file, because it embeds its trials in the HTML. That single file is
enough to reproduce the procedure in a browser.

PsychoPy (Peirce et al., 2019) is the reference target. The generated script
carries the event list as embedded JSON and interprets it, and it writes onset
triggers to the parallel port locked to the flip. It is the target to use for
EEG.

OpenSesame (Mathôt et al., 2012) is generated as a complete plain-text `.osexp`,
block by block, with each event becoming an inline script. It gets its trigger by
a different route, discussed below, and it opens the trigger device with a
fallback: if no parallel port or serial device is available, it prints the codes
and carries on, so the experiment still runs on a machine without the hardware.

jsPsych is the browser target. A browser cannot address a parallel port, so the
trigger codes go into each trial's data and no marker leaves the machine. That
makes it suitable for behavioural work, for piloting and for sharing a procedure,
though never for EEG. The generated HTML embeds the events and the trials and
saves responses locally, so no server is needed. The jsPsych library itself loads
from a CDN, so the first run needs an internet connection.

Two smaller differences are worth knowing. The OpenSesame loop is written as
`sequential` on purpose, because its default is random, which would discard the
seeded trial order and make the three targets present different sequences. The
order is worth protecting, since it is computed from the design by a keyed-hash
shuffle and is identical in the R and Python engines. And the jsPsych exporter
maps the event model's key names onto browser key names, so `left` becomes
`arrowleft`, since the same event list has to mean the same thing to a browser as
to PsychoPy.

```{r loop-table}
psychopy_csv <- file.path(out, "vignette_experiment_english_psychopy.csv")
names(read.csv(psychopy_csv))
```

The trial table carries exactly the columns the events reference plus the
bookkeeping ones. A field no event mentions is not exported, which keeps the
generated table minimal and means the trial table is itself a readable statement
of what the experiment uses.

## Trigger timing and the flip lock

This is the part that justifies generating the experiment at all. An ERP is
averaged over epochs cut around a marker, so the marker's timing error propagates
straight into the waveform. If the trigger is sent when the code decides to draw
the stimulus rather than when the display actually shows it, the error is the
interval between those two moments, which is up to one refresh (about 17 ms at
60 Hz) plus whatever the drawing code took, and it is variable from trial to
trial. A constant offset can be subtracted afterwards. Jitter cannot: it smears
the average and shrinks the component you are measuring.

The generated PsychoPy script never sends a trigger from the drawing sequence.
It registers the write with the window, to be executed on the next flip:

```
win.callOnFlip(port.setData, trigger)
```

`callOnFlip` defers the call until the flip that puts the stimulus on screen, so
the port write happens in the same operation that makes the stimulus visible.
That is the guarantee, and it is a property of the generated code. The Python
package's test suite pins it, driving the generated script against a mock PsychoPy
window and a mock port and asserting that the trigger is bound to the flip.

The reset is handled the same way. A recorder needs the line to return to zero
before the next code, so the script schedules a write of 0 on a later flip.

```{r trigger-settings}
str(schema$triggers)
```

`trigger_hold_ms` is 50, comfortably above the 10 ms minimum most recorders need
to register a pulse. The script converts it to whole flips against the refresh it
measures, so the pulse stays about 50 ms whether the display runs at 60 Hz or at
240 Hz. As a frame count it used to shorten with the refresh, reaching 12.5 ms at
240 Hz. The older `reset_after_frames` is still accepted and converted.
`inter_trigger_ms`
spaces any trailing markers. `parallel_address` is the LPT1 base address on a
typical laboratory machine, and it is a schema value because it belongs to the
machine, which will run many designs.

OpenSesame reaches the same place by a different route, and the difference is
worth being explicit about. It has no `callOnFlip`, so the generated inline script
sends the trigger on the line after `c.show()`:

```
var.onset_time = c.show()
send_trigger(var.condition_trigger)
```

This is sound because `show()` blocks until the display refresh and returns at it,
so the next statement runs immediately after the stimulus becomes visible. The
residual offset is the few microseconds of the return and the port write, and it
is constant rather than jittered. That makes it a close second to the flip lock,
which is why PsychoPy stays the reference target for EEG.

Neither generation nor its tests import PsychoPy, pyserial or any port driver.
Generation only writes text. The hardware is needed to *run* an experiment, not to
build one, which is what lets the whole demonstration reproduce on a laptop.

## Adding a paradigm

A design can override the event list outright, and this is the route that needs no
change to the package. The events below present a word twice with a mask between,
which no registered paradigm does, and both `text` events are triggered so the
recording can distinguish the two presentations.

```{r custom-events}
custom <- design
custom$name <- "vignette_custom"
custom$events <- list(
  list(type = "fixation", content = "+", duration_frames = 30L),
  list(type = "text", content = "{word}", duration_frames = 12L,
       trigger = 30L, onset_locked = TRUE),
  list(type = "mask", content = "#####", duration_frames = 6L),
  list(type = "text", content = "{word}", duration_frames = 48L,
       trigger = "condition", onset_locked = TRUE),
  list(type = "response", keys = c("left", "right"), timeout_ms = 2000L),
  list(type = "blank", duration_frames = 15L)
)
required_fields(custom)
custom_files <- export_experiments(stim, custom, schema, out)
basename(unlist(custom_files))
```

Nothing in the backends changed. All three targets rendered a sequence they had
never seen, because each one walks the event list and knows nothing about
paradigms. The design also inherits a consequence of declaring its own events: a
design with an explicit `events` list counterbalances factorially, since the
Latin-square recipe belongs to the registered paradigms that need it.

To register a paradigm for reuse across designs, add an entry to the
`PARADIGMS` registry in `R/paradigms.R`, giving its `stimulus_fields`, its
`counterbalance` recipe and its `events`, and add the matching entry to
`python_workflow/src/lexsync/paradigms.py` so that the two engines stay in step.
Designs can then name it. The registry is a plain named list and the entries above
are the pattern to copy.

Adding a presentation target is the mirror image: write an `export_<target>()`
that walks the same rendered event list, and it will serve every paradigm at once.

## References

Mathôt, S., Schreij, D., & Theeuwes, J. (2012). OpenSesame: An open-source,
graphical experiment builder for the social sciences. *Behavior Research
Methods*, *44*(2), 314–324. https://doi.org/10.3758/s13428-011-0168-7

Peirce, J., Gray, J. R., Simpson, S., MacAskill, M., Höchenberger, R., Sogo, H.,
Kastman, E., & Lindeløv, J. K. (2019). PsychoPy2: Experiments in behavior made
easy. *Behavior Research Methods*, *51*(1), 195–203.
https://doi.org/10.3758/s13428-018-01193-y
