---
title: "The app"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The app}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "",
  eval = FALSE,
  R.options = list(
    cli.num_colors = 1,
    cli.hyperlink = FALSE,
    crayon.enabled = FALSE,
    width = 80
  )
)
# Console colour carries no meaning on a rendered page. pkgdown turns it on for
# its own build, and the escape sequences then reach the reader as literal text,
# so colour is switched off here for a plain vignette render and a site build
# alike. The fixed width keeps printed output inside the documentation column.
```

Not every design has to be written as code. The repository carries a Shiny
front-end over this package which drives the whole workflow through a browser
tab, from choosing a paradigm to downloading the stimuli, the realised-control
report and the materials datasheet. It is a front-end and nothing more. Every
number it shows comes from the installed `lexsync` package, never from a
re-implementation, and the design it assembles is an ordinary YAML file that you
can save, share and run from a script afterwards.

This article describes the app without running it, since the app needs a live
server, so the code below is shown but not evaluated.

## Launching it

The app is not part of the package. It lives in the repository, under
`apps/r_shiny`, beside the corpora and the example item tables that it reads, so
launch it from a clone. The interface needs shiny, bslib and DT, and the
download button needs zip. All four are listed in the package's `Suggests`.

```{r}
install.packages(c("shiny", "bslib", "DT", "zip"))
shiny::runApp("apps/r_shiny", port = 8502)
```

Run that from the root of the clone. The app looks for `corpora/derived/` in the
working directory and then in the two directories above it, and it finds the
example item tables under `items/` the same way, so a launch from somewhere else
leaves it with no lexicon to offer. It opens at `http://127.0.0.1:8502`.

The repository also carries a launcher for the terminal, which takes an optional
port and is what a container or a headless session should use.

```sh
Rscript apps/r_shiny/run.R          # 8502 unless a port is given
Rscript apps/r_shiny/run.R 8600
```

The launcher fixes the host and the port and never opens a browser itself. A
server started with an auto-launched browser or an unbound port can exit the
moment it is backgrounded, which is the usual reason an app appears not to stay
up in an automated environment.

## The sidebar

Whatever the paradigm, the same few fields sit in the sidebar. 'Paradigm'
chooses between a factorial word contrast drawn from a corpus, lexical decision
with generated pseudowords, priming and self-paced reading, the last two reading
prepared items from a table. 'Design name' and 'Language label' become the slug
that every written file is named after, exactly as in a design file, so
`my_design` plus `english` gives `my_design_english_stimuli_R.csv` and its
companions. 'Stimulus font' overrides the presentation font, which matters for a
non-Latin script, and it reaches the design only when it differs from the
`Courier New` default, leaving the schema in charge otherwise.

Under those the sidebar prints the version of lexsync that is running, with the
reminder that the app runs the installed package. If that line disagrees with the
version you expect, the app is reading a different install.

## Describing the design

The main panel changes with the paradigm, since a corpus-matched contrast and a
prepared item table need different things.

### A factorial word contrast

'Lexicon' lists the derived corpora the clone holds under `corpora/derived/`,
which are English, Spanish, a Spanish lexicon carrying grammatical gender, and
Mandarin Chinese. Two sliders then narrow the lexicon to the candidate pool the
design will consider at all, one for length in letters or characters and one for
frequency in Zipf. They are the design's `pool_filters`, and they are a step that
matters, because matching never re-reads them.

The conditions are an editable table. Each row names a condition and carves it
out of the pool, either by a numeric window through the `dimension`, `lower` and
`upper` columns or by a set of comma-separated values through `categories`, which
is what a categorical column such as the gender of the Spanish lexicon needs. The
optional `dimension2`, `lower2` and `upper2` columns add a second factor to the
same row, so one row can define a full cell of a 2 × 2. Three presets fill the
table in one click, a high versus low frequency contrast, a dense versus sparse
neighbourhood contrast and a 2 × 2 of frequency by neighbourhood, and every cell
stays editable afterwards, so a preset is only a starting point.

Below the table sit the choices that govern the match itself. 'Items per
condition' sets how many stimuli each condition should hold. 'Matching method'
offers the four the engine implements, `standardised_euclidean`, `joint`,
`mahalanobis` and `optimal`. 'Resampled disjoint sets' draws several matched sets
that share no item, which is what lets items be treated as a random factor, and
zero switches it off. 'Match on' picks the dimensions to equate across
conditions, from length, frequency, neighbourhood N, OLD20, syllable count and
bigram frequency. 'Counterbalancing lists' sets how many lists the items are
rotated across.

The advanced panel underneath holds one tolerance window per matched dimension,
the half-width in standard deviations of the band a candidate must fall inside.
It is closed by default because the schema already carries a sensible default for
every dimension, and a zero there leaves that default alone, where you might expect
it to pin the window to nothing. Reproducing the exact window of a published study
is the usual reason to touch the panel at all, and the panel itself gives
the worked example that *Matching, dimensions and designs* discusses, a frequency
k of 0.111 for a band of the mean plus or minus a ninth of a standard deviation.

### Lexical decision

A lexical-decision design takes the same lexicon and pool filters and then asks
only for the number of items per condition, words and pseudowords being equal in
number, and for the generator. `letter_substitution`, the engine default, changes
as few single letters as it can while keeping every bigram attested.
`subsyllabic` swaps whole onset, nucleus and coda constituents in the manner of
Wuggy (Keuleers & Brysbaert, 2010). Both are deterministic, so the pseudowords
are the same ones the package would generate from a script.

### Priming and self-paced reading

These two paradigms read prepared items instead of selecting any, so no corpus
and no matching are involved. The app uses the bundled example table for the
paradigm, `items/priming_pairs_en.csv` or `items/spr_sentences_en.csv`, and asks
only how many counterbalancing lists to build.

## Running it

'Run design' assembles the design, writes it to a `design.yaml` in a temporary
directory with the lexicon and item paths resolved to absolute ones, and hands it
to `run_pipeline()` with the schema that ships inside the package. Two conditions
are checked first, that a lexicon has been chosen where the paradigm needs one
and that a factorial design defines at least two conditions, and anything the
pipeline itself refuses comes back as the error the pipeline raised, in its own
words. A successful run reports how many rows were selected and opens the results.

## Reading the results

The results arrive as six tabs, which between them cover what was selected,
whether it is any good, what it will present and how to reproduce it.

'Stimuli' is the selected set as a table, the same one the run wrote to disk.

'Realised control' is the part that decides whether the set is usable. It reports
Cohen's *d* against the anchor condition for each controlled dimension, with its
90% confidence interval, the variance ratio and the verdict of the two one-sided
tests procedure, then draws the absolute standardised mean difference per
dimension as a bar chart with a dashed line at 0.5 standard deviations.
Manipulated dimensions stand well above that line and matched ones sit near zero,
so the chart shows immediately whether the matching worked. The per-condition
descriptive statistics follow underneath. A design drawn from an item table
produces no corpus-matching report, and the tab says as much where an empty table
would otherwise sit.

'Datasheet' renders the materials datasheet for the run, carrying the provenance
of the corpus, the checksums, the realised control and the pre-registration
skeleton.

'Experiment scripts' offers one download per generated experiment, the PsychoPy
script, the OpenSesame `.osexp` and the jsPsych HTML that runs in a browser, all
three compiled from the same trial description.

'Reproducible code' is what stops a session in the app from being a dead end. It
shows the design configuration as YAML, with repository-relative paths such as
`corpora/derived/en.csv` rather than the temporary ones the run used, followed
by the one-line R, Python and command-line calls that reproduce the same
operation. Saving the YAML under the name the panel gives it and running any of
those three lines repeats the run outside the app.

'Download' packages the design and every artefact the run wrote, the stimuli, the
reports, the datasheet, the run log and the three experiments, into a single
archive named after the design.

## What you leave with

A session in the app ends with the same artefacts a scripted run produces, which
is why it was built as a front-end over the package and not as a tool of its own.
The design YAML is portable across both engines, so a set assembled here can be
reproduced by a collaborator working in Python, byte for byte under the
deterministic matching methods. *Reproducibility, parity and the materials
datasheet* covers what that guarantee promises and where it stops, and
*Matching, dimensions and designs* reads the realised-control report column by
column.
